---
title: "Penda Health"
author: "Pius Mbeti"
date: "2023-07-23"
output:
  pdf_document: default
  html_document: default
---



**INTRODUCTION** \
In this report, I present an analysis of data from Penda Health Clinic. The data includes information on medical visits, revenue, and diagnoses recorded at the clinic. We will use R for data processing, Exploratory Data Analysis (EDA), and analysis, and then create visualizations to gain insights into the clinic's performance. \

**DATA LOADING AND OVERVIEW** \
Before the analysis, let's first take a look at the structure of the data.  




```{r}
setwd("C:/Users/Pius/OneDrive/Desktop/Penda health Project")



# Set warning behavior to ignore
options(warn = -1)  # -1 suppresses all warnings






# Read the CSV files into data frames
visit_tbl <- read.csv("Visit_Tbl.csv")
invoice_tbl <- read.csv("Invoice_Tbl.csv")
diagnosis_tbl <- read.csv("Diagnosis_Tbl.csv")

# Merging the three tableS
merged_data <- merge(visit_tbl, invoice_tbl, by = "VisitCode", all.x = TRUE)
merged_data <- merge(merged_data, diagnosis_tbl, by = "VisitCode", all.x = TRUE)
head(merged_data)
str(merged_data)


```

**Exploratory Data Analysis (EDA)**\
Exploring the data and understanding its characteristics. \
Summary statistics for merged_data

```{R}

summary(merged_data)



```


*Explore the distribution of billed amount*

```{R}

library(ggplot2)

ggplot(merged_data, aes(x = Amount)) +
  geom_histogram(binwidth = 500, fill = "#1f77b4") +
  labs(title = "Distribution of Billed Amount",
       x = "Billed Amount",
       y = "Frequency") +
  theme_minimal()

```


*Explore the distribution of diagnoses*
```{r}
 
ggplot(merged_data, aes(x = Diagnosis)) +
  geom_bar(fill = "#1f77b4") +
  labs(title = "Distribution of Diagnoses",
       x = "Diagnosis",
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))



```

**Data Processing and Wrangling** \
Before creating visualizations, I need to process and clean the data to ensure it's in the right format.

```{r}
# Convert "VisitDateTime" to a proper date-time format
merged_data$VisitDateTime <- as.POSIXct(merged_data$VisitDateTime, format = "%m/%d/%Y %H:%M")

# Convert "NPS.Score" to numeric and handle missing values
merged_data$NPS.Score <- as.numeric(merged_data$NPS.Score)

# Convert "Amount" to numeric
merged_data$Amount <- as.numeric(merged_data$Amount)

# Handle missing values in the "Diagnosis" column
merged_data$Diagnosis[is.na(merged_data$Diagnosis)] <- "Unknown"

head(merged_data)
str(merged_data)

```

**DATA VISUALIZATION** \\
*Key Performance Indicators* \
Calculating and displaying some key performance indicators for the clinic.

```{r}
# Load required libraries

library(dplyr)
# Calculating KPIs
total_visits <- sum(merged_data$VisitCategory == "In-person Visit" | merged_data$VisitCategory == "Telemedicine Visit")
total_revenue <- sum(merged_data$Amount)
total_patients <- n_distinct(merged_data$PatientCode)

# Display KPIs
cat("Total Visits: ", total_visits, "\n")
cat("Total Revenue: Sh.", total_revenue, "\n")
cat("Total Patients: ", total_patients, "\n")



```

**Monthly Trends of Number of Visits**

```{r}
# Data preparation for monthly trends
merged_data$VisitDateTime <- as.Date(merged_data$VisitDateTime, format = "%m/%d/%Y %H:%M")
monthly_visits <- merged_data %>%
  filter(format(VisitDateTime, "%b") %in% c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")) %>%
  group_by(Month = factor(format(VisitDateTime, "%b %Y"), levels = c("Mar 2022", "Apr 2022", "May 2022", "Jun 2022", "Jul 2022", "Aug 2022", "Sep 2022", "Oct 2022"))) %>%
  summarise(Visits = n())

# Visualization: Line Graph
ggplot(monthly_visits, aes(x = Month, y = Visits, group = 1)) +
  geom_line(color = "#1f77b4", size = 1.5) +
  geom_point(color = "#1f77b4", size = 3) +
  labs(title = "Monthly Trends of Number of Visits",
       x = "Month",
       y = "Number of Visits") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
The chart revealed that the month of June recorded the highest number of visits, while the month of October had the lowest. This insight allows the clinic to better prepare for peak months and efficiently manage patient flow. \ \


**Number of Visits per Medical Center**
```{r}
# Data preparation for visits per medical center
visits_per_center <- merged_data %>%
  group_by(MedicalCenter) %>%
  summarise(Visits = n()) %>%
  arrange(desc(Visits))

# Visualization: Bar Chart
ggplot(visits_per_center, aes(x = reorder(MedicalCenter, -Visits), y = Visits, fill = MedicalCenter)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Visits per Medical Center",
       x = "Medical Center",
       y = "Number of Visits") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The bar chart comparing the number of visits per medical center highlighted "Tassia" as the most visited center, with "Mathare North" recording the least number of visits. This insight provides actionable information to optimize service distribution and improve underperforming centers. \ \

**Revenue Generated by Medical Centers**
```{r}

# Data preparation for revenue per medical center
revenue_per_center <- merged_data %>%
  group_by(MedicalCenter) %>%
  summarise(TotalRevenue = sum(Amount)) %>%
  arrange(desc(TotalRevenue))

# Visualization: Stacked Bar Chart with revenue in millions
ggplot(revenue_per_center, aes(x = reorder(MedicalCenter, -TotalRevenue), y = TotalRevenue, fill = MedicalCenter)) +
  geom_bar(stat = "identity") +
  labs(title = "Revenue Generated by Medical Centers",
       x = "Medical Center",
       y = "Total Revenue (Millions)") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +  # Format y-axis labels in millions
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()


```
The chart showcasing revenue generated by medical centers identified "Tassia" as the highest revenue-generating center, while "Call Centre" contributed the least. This insight guides revenue enhancement strategies, focusing on high-revenue centers to further boost overall revenue. \ \


**Diagnoses and Total Patients Recorded**

```{r}

# Data preparation for diagnoses and total patients recorded
diagnoses_table <- merged_data %>%
  group_by(Diagnosis) %>%
  summarise(TotalPatients = n()) %>%
  arrange(desc(TotalPatients))

# Display the matrix table
pander::pander(diagnoses_table, caption = "Diagnoses and Total Patients Recorded")



```
The table shows the total patients recorded for each diagnosis. "Acute Nasopharyngitis" emerged as the most common diagnosis with 4441 patients, allowing the clinic to prioritize and allocate resources for managing prevalent health conditions. \ \





**CONCLUSION** \
In conclusion, this data analysis report provides valuable insights into the performance of Penda Health Clinic. I have performed Exploratory Data Analysis (EDA) to understand the data distribution and then analyzed various key performance indicators, monthly visit trends, number of visits per medical center, revenue generated by medical centers, and diagnoses with the highest number of patients recorded. These insights can help the clinic in making data-driven decisions for improved performance and patient care. \ \



**RECOMMENDATIONS** \
*1. Resource Optimization:* Allocate resources based on monthly visit trends to efficiently handle higher patient volumes during peak months like June. This will improve patient experience and reduce wait times. \

*2. Focus on High-Revenue Centers:* Implement targeted marketing and service enhancement strategies for medical centers like "Tassia," which generate higher revenue. This will further maximize revenue generation for the clinic. \

*3. Enhanced Diagnoses-Specific Care:* Given the prevalence of "Acute Nasopharyngitis," develop specialized care plans and resources to manage and treat this condition effectively. Proactive measures can improve patient outcomes and satisfaction. \

*4. Data-Driven Decision Making:* Continue utilizing data analytics and interactive dashboards to drive informed decision-making in areas such as resource allocation, revenue enhancement, and patient care.

